|

### 2023-03-09 paper/kaggle.Rmd
more clean-up; moved Kaggle nb to separate file; started intro

---
title: |
  Introducing STRAIN: A simple and interpretable pass rush metric inspired by materials science
short_title: STRAIN
authors: 
- name: "Author 1"
  affiliation: "Address"
  city: "City"
  country: "Country"
  email: "email1\\@example.com"
- name: "Author 2"
  affiliation: "Address2"
  city: "City2"
  country: "Country2"
  email: "email2\\@example.com"
abstract: |
  Abstract goes here
keywords: tracking data, Bornn legacy, too much Canadian
bibliography: "bibliography"
output: rticles::rss_article
---

\newpage

**PAPER SKETCH**

i. Introduction

<!--
TODO: update citations after they're published
-->

Tracking data in recent years, NFL

examples of past nfl tracking data projects

See @Baumer2023Big, @Macdonald2020Recreating, and @Kovalchik2023Player for more details

Importance of D-linemen, pass rush (Maybe talk about role of each position, edge vs interior, defensive formation)

Previous metrics: Greg's lit review, their flaws

Our contribution

ii. Methods

a. data

2023 bdb [@bdb2023]

first 8 weeks 2021 season

tracking: locations of players and ball, 10 fps,

pre-processing standardize

PFF scouting: pass rush role

sample restriction: remove plays w multiple QB

frames between snap and QB outcome (pass forward, sack)

sample size

b. strain rate

c. applying to pass rush

iii. Results

example play

positional curves

rankings (indiv, add more for team-level offense/defense)

corr with pressure rate = (hits + sacks + hurries) per snap

meta metrics: discriminant, stability, independence @Franks2016Meta

iv. Discussion

\newpage

@@

## Previous Metrics for Pass Rush Evaluation

## Our Contribution

<!--

Greg's long lit review

The theme of Big Data Bowl 2023 is about evaluating performance of linemen, both offensive and defensive, on passing plays. While the ideas presented here could be extended to offensive linemen, we have chosen to focus on measuring the performance of defensive linemen. Officially tracked by the NFL since 1982, the most well known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass. Other counting statistics have been tracked by various entities such as hurries and hits as well as pressures, which is defined as the sum of hurries, hits, and sacks. More recently metrics such as Time In Pocket (TIP) and Pass Rusher Win Rate (PRWR) have also been proposed and are substantial improvements over the less sophisticated counting statistics. While PRWR is certainly a step in the right direction, it takes continuous data and converts it to a binary win/loss, which is generally not recommended in statistical analysis, thus, the metric we propose here is computed continuously over time just as the data is actually observed.

The metric we propose is based on the following simple question: What is considered a success for a pass rusher on a pass play? The simplest answer to this question, and the first defensive lineman statistics to be tracked by the NFL, is a sack. In terms of tracking data, one can think of a sack as reducing the distance between the pass rusher and the quarterback to 0. Therefore, a decent starting point for a continuous measure of pass rushing effectiveness shoould be based on how close the pass rusher is able to get to the quarterback with smaller distances indicating amore successful pass rush. However, a good measure of pass rushing effectiveness cannot simply be based on distance to the quarterback; Consider the following case: a pass rusher gets within two yards of the quarterback, but is then stopped and fails to get any closer for the rest of the play. This is not a successful rush because the pass rusher is not continuing to reduce the distance between themselves and the quarterback. Therefore, a good metric should consider the rate at which this distance is being reduced. However, just like distance, rate by itself is inadequate in assessing success of a pass rusher. Consider the following case: A defensive player is 50 yards away and moving towards the quarterback very rapidly. Their moving rate is essentially unimportant due to a significantly far distance away from the quarterback.

This means that in order to appropriately evaluate rusher performance, the distance from the quarterback and the rate at which this distance is being reduced must both be considered. Therefore, we propose STRAIN, a new measure for pass rusher effectiveness, which is computed as the ratio of the rate that the pass rusher is reducing the distance between themselve and the quarterback and distance between the pass rusher and the quarterback. In performing some research about this proposed quantity, we discovered that this concept actually already exists in materials science and is called the "strain rate" (which is where the name of our metric is derived). Given this connection, one can view the players as particles in some material and the defensive "particles" are attempting to exert pressure on the pocket in an attempt to compress and collapse this pocket around the quarterback. We think this analogy is just delightful.

@@

-->

The remainder of this paper is outlined as follows.

@@

## Data

bdb

@@

* The distance between player $j$ and the quarterback at frame $t$ during play $i$ is

$$f_{d_{ij}}(t) = \sqrt{\left(x_{ijt} - x^{QB}_{it}\right)^2 + \left(y_{ijt} - y^{QB}_{it}\right)^2}$$.

* The rate at which player $j$ is moving towards the quarterback at frame $t$ during play $i$ is

$$\displaystyle f'_{d_{ij}}(t) = \frac{df_{d_{ij}}(t)}{dt}$$.

* The STRAIN for player $j$ at frame $t$ during play $i$ is

@@

# Results

@@

<!--

To determine the most effective pass rushers, we aggregate across all frames played and calculate the expected STRAIN (xSTRAIN) for each player, which is computed by simply averaging the STRAIN across all the frames of a play. The following tables show the top performing pass rushers (with at least 100 plays) rated by xSTRAIN for the first 8 weeks of the 2021 season.

The top interior rusher for the first 8 weeks of the 2021 season according to our xSTRAIN metric was Aaron Donald, almost universally considered the best interior rusher in the NFL currently with an xSTRAIN of 1.67. Through the first 8 weeks of the 2021 season, Donald had an impressive 5 sacks and 12 quarterback hits. Donald Was followed by two Raiders: Solomon Thomas and Quinton Jefferson with xSTRAIN values of 1.51 and 1.46, respectively. Rounding out the top 5 were Chris Jones with an xSTRAIN of 1.42 and DeForest Buckner with an xSTRAIN of 1.26.

As for edge rushers, Rashan Gary with an xSTRAIN value of 2.82 is the top of our list for the first 8 weeks of the 2021 season. Over this time period, Gary had 3.5 sacks and 15 quarterback hits. Second on our list of edge rushers is Leonard Floyd with an xSTRAIN value of 2.80 who had 6.5 sacks and 9 quarterback hits over the time period observed. Rounding out the top 5 are Justin Houston (2.78), Myles Garrett (2.75), and Von Miller (2.75). The sixth player on our rankings of edge rushers over the first 8 weeks on the 2021 season is the 2021 Defensive Player of the Year, T.J. Watt (2.71) who had 11.5 sacks and 18 quarterback hits in those first 8 weeks. Notably their is considerable overlap between our top 20 list over the first 8 weeks of the 2021 season and the Pro Football Focus edge rusher rankings prior to the 2022 season.

-->

To determine the most effective pass-rushers, we compute $\overline{\text{STRAIN}}$, which represents the average STRAIN across all frames played for each player.
Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass-rushers (NT's and DT's) separately from edge rushers (OLB's and DE's).
The following tables show the top performing pass-rushers (with at least 100 plays) rated by $\overline{\text{STRAIN}}$ for the first 8 weeks of the 2021 season.

@@

<!--

We have proposed a simple, effective, and easily interpretable metric for evaluating the performance of pass rushers in the NFL that we call STRAIN. Higher values of STRAIN indicate two important aspects of a pass rush: 1) the distance between the pass rusher and the quarterback is small and 2) the rate at which the pass rusher is moving towards the quarterback is high.

One of the major advantages of STRAIN over existing measures is that it gives a measure of effectiveness of a pass rush for every single play at a much more granular level than simply the play resulted in a sack or not or the play was a "success" or not. It is important to have a quantitative measure of pass rushing effectiveness if one wants to objectively evaluate the performance of pass rushers. STRAIN is just such a measure that objectively measures the pressure that pass rushers put on a quarterback.

In this report, we simply propose the new metric STRAIN; However, there are innumerable future projects that could be based on STRAIN. For example, STRAIN could be used to collectively evaluate the performance of a pass rushing unit by looking at the total amount of STRAIN during a play, which could, in turn, be used to evaluate the effectiveness of different types of blitzes. In addtion, STRAIN could be included as a covariate in models trying to predict quantities such as passing yards or completion probability. Notably, I believe there is very interesting work that could be done with functional regression (i.e. treating the STRAIN curve over time as a function) in modeling how STRAIN affects offensive production though is non-trivial as the length in time of plays varies from one play to another, and we leave this as future work.

In addtion, while we focused solely on evaluating the performance of pass rushers, this concept can be easily extended to the evaluation of pass blockers as a unit by simply looking at a quantity such as the total amount of STRAIN on a play or the maximum amount of STRAIN per frame aggregated across the entire play. Further, STRAIN could be extended to individual offensive pass blockers provided that a method of matching blockers to rushers was devised.

While STRAIN is a big step forward, there are still some notable limitations to this metric. For instance, STRAIN, as currently devised, cannot distinguish between a pass rusher who has to face a single blocker versus a double team. This problem could be addressed, at least somewhat, by devising a method of matching pass rushers to blockers as previously mentioned. Another limitation of this metric occurs when a quarterback scrambles out of the pocket. If you are a pass rusher on the left side and the quarterback scrambles out to the left, the STRAIN value for that pass rusher can drop considerably as the distance between the quarterback and the pass rusher gets larger, which the pass rushers shouldn't necessarily be negatively affected by in a measure of pass rushing effectiveness.

Finally, we believe there are other concepts in materials science that could be applied to evaluating pass rushing and pass blocking. One potential idea is a quantity called stress, which measures force over an area. Recall from intro physics that force is mass times acceleration, so this quantity would take into account the physical size of a pass rusher in the computation of force. You could then somehow divide this force over the "area" of the pocket formed by pass blockers to compute stress. There is also a quantity measuring the ratio of stress and strain called the elastic modulus which measures "an object's or substance's resistance to being deformed elastically". It is easy to imagine the pocket as an object and evaluating the succes of the pocket based on it's "resistance to being deformed".

-->

>>>>

---
output: html_document
---

@@

<!--
Greg's long lit review
The theme of Big Data Bowl 2023 is about evaluating performance of linemen, both offensive and defensive, on passing plays. While the ideas presented here could be extended to offensive linemen, we have chosen to focus on measuring the performance of defensive linemen. Officially tracked by the NFL since 1982, the most well known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass. Other counting statistics have been tracked by various entities such as hurries and hits as well as pressures, which is defined as the sum of hurries, hits, and sacks. More recently metrics such as Time In Pocket (TIP) and Pass Rusher Win Rate (PRWR) have also been proposed and are substantial improvements over the less sophisticated counting statistics. While PRWR is certainly a step in the right direction, it takes continuous data and converts it to a binary win/loss, which is generally not recommended in statistical analysis, thus, the metric we propose here is computed continuously over time just as the data is actually observed.
The metric we propose is based on the following simple question: What is considered a success for a pass rusher on a pass play? The simplest answer to this question, and the first defensive lineman statistics to be tracked by the NFL, is a sack. In terms of tracking data, one can think of a sack as reducing the distance between the pass rusher and the quarterback to 0. Therefore, a decent starting point for a continuous measure of pass rushing effectiveness shoould be based on how close the pass rusher is able to get to the quarterback with smaller distances indicating amore successful pass rush. However, a good measure of pass rushing effectiveness cannot simply be based on distance to the quarterback; Consider the following case: a pass rusher gets within two yards of the quarterback, but is then stopped and fails to get any closer for the rest of the play. This is not a successful rush because the pass rusher is not continuing to reduce the distance between themselves and the quarterback. Therefore, a good metric should consider the rate at which this distance is being reduced. However, just like distance, rate by itself is inadequate in assessing success of a pass rusher. Consider the following case: A defensive player is 50 yards away and moving towards the quarterback very rapidly. Their moving rate is essentially unimportant due to a significantly far distance away from the quarterback.
This means that in order to appropriately evaluate rusher performance, the distance from the quarterback and the rate at which this distance is being reduced must both be considered. Therefore, we propose STRAIN, a new measure for pass rusher effectiveness, which is computed as the ratio of the rate that the pass rusher is reducing the distance between themselve and the quarterback and distance between the pass rusher and the quarterback. In performing some research about this proposed quantity, we discovered that this concept actually already exists in materials science and is called the "strain rate" (which is where the name of our metric is derived). Given this connection, one can view the players as particles in some material and the defensive "particles" are attempting to exert pressure on the pocket in an attempt to compress and collapse this pocket around the quarterback. We think this analogy is just delightful.
-->

@@

@@

@@

* The distance between player $j$ and the quarterback at frame $t$ during play $i$ is $f_{d_{ij}}(t) = \sqrt{(x_{ijt} - x^{QB}_{it})^2 + (y_{ijt} - y^{QB}_{it})^2}$.
* The rate at which player $j$ is moving towards the quarterback at frame $t$ during play $i$ is $\displaystyle f'_{d_{ij}}(t) = \frac{df_{d_{ij}}(t)}{dt}$.
* The STRAIN for player $j$ at frame $t$ during play $i$ is

@@

# Analysis

@@

<!--
To determine the most effective pass rushers, we aggregate across all frames played and calculate the expected STRAIN (xSTRAIN) for each player, which is computed by simply averaging the STRAIN across all the frames of a play. The following tables show the top performing pass rushers (with at least 100 plays) rated by xSTRAIN for the first 8 weeks of the 2021 season.
The top interior rusher for the first 8 weeks of the 2021 season according to our xSTRAIN metric was Aaron Donald, almost universally considered the best interior rusher in the NFL currently with an xSTRAIN of 1.67. Through the first 8 weeks of the 2021 season, Donald had an impressive 5 sacks and 12 quarterback hits. Donald Was followed by two Raiders: Solomon Thomas and Quinton Jefferson with xSTRAIN values of 1.51 and 1.46, respectively. Rounding out the top 5 were Chris Jones with an xSTRAIN of 1.42 and DeForest Buckner with an xSTRAIN of 1.26.
As for edge rushers, Rashan Gary with an xSTRAIN value of 2.82 is the top of our list for the first 8 weeks of the 2021 season. Over this time period, Gary had 3.5 sacks and 15 quarterback hits. Second on our list of edge rushers is Leonard Floyd with an xSTRAIN value of 2.80 who had 6.5 sacks and 9 quarterback hits over the time period observed. Rounding out the top 5 are Justin Houston (2.78), Myles Garrett (2.75), and Von Miller (2.75). The sixth player on our rankings of edge rushers over the first 8 weeks on the 2021 season is the 2021 Defensive Player of the Year, T.J. Watt (2.71) who had 11.5 sacks and 18 quarterback hits in those first 8 weeks. Notably their is considerable overlap between our top 20 list over the first 8 weeks of the 2021 season and the Pro Football Focus edge rusher rankings prior to the 2022 season.
-->

To determine the most effective pass-rushers, we compute <span style="text-decoration:overline; font-weight:bold">STRAIN</span>, which represents the average STRAIN across all frames played for each player.
Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass-rushers (NT's and DT's) separately from edge rushers (OLB's and DE's).
The following tables show the top performing pass-rushers (with at least 100 plays) rated by <span style="text-decoration:overline; font-weight:bold">STRAIN</span> for the first 8 weeks of the 2021 season.

@@

<!--
We have proposed a simple, effective, and easily interpretable metric for evaluating the performance of pass rushers in the NFL that we call STRAIN. Higher values of STRAIN indicate two important aspects of a pass rush: 1) the distance between the pass rusher and the quarterback is small and 2) the rate at which the pass rusher is moving towards the quarterback is high.
One of the major advantages of STRAIN over existing measures is that it gives a measure of effectiveness of a pass rush for every single play at a much more granular level than simply the play resulted in a sack or not or the play was a "success" or not. It is important to have a quantitative measure of pass rushing effectiveness if one wants to objectively evaluate the performance of pass rushers. STRAIN is just such a measure that objectively measures the pressure that pass rushers put on a quarterback.
In this report, we simply propose the new metric STRAIN; However, there are innumerable future projects that could be based on STRAIN. For example, STRAIN could be used to collectively evaluate the performance of a pass rushing unit by looking at the total amount of STRAIN during a play, which could, in turn, be used to evaluate the effectiveness of different types of blitzes. In addtion, STRAIN could be included as a covariate in models trying to predict quantities such as passing yards or completion probability. Notably, I believe there is very interesting work that could be done with functional regression (i.e. treating the STRAIN curve over time as a function) in modeling how STRAIN affects offensive production though is non-trivial as the length in time of plays varies from one play to another, and we leave this as future work.
In addtion, while we focused solely on evaluating the performance of pass rushers, this concept can be easily extended to the evaluation of pass blockers as a unit by simply looking at a quantity such as the total amount of STRAIN on a play or the maximum amount of STRAIN per frame aggregated across the entire play. Further, STRAIN could be extended to individual offensive pass blockers provided that a method of matching blockers to rushers was devised.
While STRAIN is a big step forward, there are still some notable limitations to this metric. For instance, STRAIN, as currently devised, cannot distinguish between a pass rusher who has to face a single blocker versus a double team. This problem could be addressed, at least somewhat, by devising a method of matching pass rushers to blockers as previously mentioned. Another limitation of this metric occurs when a quarterback scrambles out of the pocket. If you are a pass rusher on the left side and the quarterback scrambles out to the left, the STRAIN value for that pass rusher can drop considerably as the distance between the quarterback and the pass rusher gets larger, which the pass rushers shouldn't necessarily be negatively affected by in a measure of pass rushing effectiveness.
Finally, we believe there are other concepts in materials science that could be applied to evaluating pass rushing and pass blocking. One potential idea is a quantity called stress, which measures force over an area. Recall from intro physics that force is mass times acceleration, so this quantity would take into account the physical size of a pass rusher in the computation of force. You could then somehow divide this force over the "area" of the pocket formed by pass blockers to compute stress. There is also a quantity measuring the ratio of stress and strain called the elastic modulus which measures "an object's or substance's resistance to being deformed elastically". It is easy to imagine the pocket as an object and evaluating the succes of the pocket based on it's "resistance to being deformed".
-->

### 2023-03-09 paper/paper.Rmd
more clean-up; moved Kaggle nb to separate file; started intro

---
title: |
  Introducing STRAIN: A simple and interpretable pass rush metric inspired by materials science
short_title: STRAIN
authors: 
- name: "Author 1"
  affiliation: "Address"
  city: "City"
  country: "Country"
  email: "email1\\@example.com"
- name: "Author 2"
  affiliation: "Address2"
  city: "City2"
  country: "Country2"
  email: "email2\\@example.com"
abstract: |
  Abstract goes here
keywords: tracking data, Bornn legacy
bibliography: "bibliography"
output: rticles::rss_article
---

@@

A paragraph on tracking data

@@

See @Baumer2023Big, @Macdonald2020Recreating, and @Kovalchik2023Player for more details

Importance of linemen

@@

Our contribution

# Data

@@

@@

example play

positional curves

rankings (indiv, add more for team-level offense/defense)

corr with pressure rate = (hits + sacks + hurries) per snap

meta metrics: discriminant, stability, independence

@Franks2016Meta

>>>>

---
title: |
  Introducing STRAIN: A simple and interpretable pass rush metric inspired by materials science
short_title: STRAIN
authors: 
- name: "Author 1"
  affiliation: "Address"
  city: "City"
  country: "Country"
  email: "email1\\@example.com"
- name: "Author 2"
  affiliation: "Address2"
  city: "City2"
  country: "Country2"
  email: "email2\\@example.com"
abstract: |
  Abstract goes here
keywords: tracking data, Bornn legacy, too much Canadian
bibliography: "bibliography"
output: rticles::rss_article
header-includes:
  \usepackage{setspace}
  \usepackage{booktabs}
---

\newpage

@@

@@

In recent years, tracking data have replaced traditional box-score statistics and play-by-play data as the state-of-the-art tool for analyzing and deeper understanding different sports.

examples of past nfl tracking data projects

See @Baumer2023Big, @Macdonald2020Recreating, and @Kovalchik2023Player for more complete reviews on how tracking data have transformed sports analytics.

NFL, next gen stats https://operations.nfl.com/gameday/technology/nfl-next-gen-stats/

Importance of D-linemen, pass rush (Maybe talk about role of each position, edge vs interior, defensive formation)

@@

## Previous Pass Rush Metrics

\begin{table}
\caption{A summary of previously-existed pass rush metrics. \label{tab:metrics}}
\centering
\begin{tabular}{ll}
\toprule
\textbf{Metric} & \textbf{Description}\\
\midrule
Sacks & a\\
Hits & a\\
Hurries & a\\
QB Pressures & a\\
Pass-Rush Productivity (PRP) & a\\
Time In Pocket (TIP) & a\\
Pass Rush Win Rate (PRWR) & Rate at which pass rusher beats pass block within 2.5 seconds after snap\\
\bottomrule
\end{tabular}
\end{table}

Table \ref{tab:metrics}

Sacks, hits, hurries, pressures

Officially tracked by the NFL since 1982, the most well known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.

https://www.pro-football-reference.com/about/advanced_stats.htm

Pressures: QB pressures (hurries + knockdowns + sack plays). Note that "sack plays" in this instance include both sacks and assists.

Hurried: Times QBs were hurried - defined as a situation when the QB is forced to throw the ball earlier than intended or is chased around/out of the pocket as the result of defensive pressure

Hit: Times QBs were hit - occurs when the QB hits the ground after a throw, not marked on a sack play

PFF's Pass-Rush Productivity (PRP)

https://www.pff.com/news/pro-signature-stats-spotlight-defensive-line

More recently metrics such as Time In Pocket (TIP) and Pass Rush Win Rate (PRWR) have also been proposed and are substantial improvements over the less sophisticated counting statistics.

Pocket Time: Number of seconds the quarterback was in the pocket between the snap of the ball and a pass or the collapse of the pocket

@Burke2018Created

https://www.espn.com/nfl/story/_/id/24892208

2.5 seconds, arbitrary

While PRWR is certainly a step in the right direction, it takes continuous data and converts it to a binary win/loss, which is generally not recommended in statistical analysis

## Contributions of STRAIN

The remainder of this paper is outlined as follows.

@@

## Data

2023 bdb [@Howard2023NFL]

first 8 weeks 2021 season

tracking: locations of players and ball, 10 fps,

pre-processing standardize

PFF scouting: pass rush role

sample restriction: remove plays w multiple QB

frames between snap and QB outcome (pass forward, sack)

sample size

## Strain rate in materials science

## Application to pass-rushing in football

@@

## Example play

## Positional STRAIN curves

## Ranking the best pass-rushers

## Corr with pressure rate = (hits + sacks + hurries) per snap

## Statistical Properties of STRAIN

From @Franks2016Meta (github https://github.com/afranks86/meta-analytics)

Discrimination: Does the metric reliably differentiate between players?

Stability: Does the metric measure a quantity which is stable over time?

Independence: Does the metric provide new information?

# Discussion

# References
